---
title: "Forecasting Stock Prices Using ARIMA and Benchmark Methods"
author:
  - Catherine Lu
  - Emily Wang
  - Katrina Wang
date: May 7, 2026
format: pdf
---

```{r, include=FALSE}
#packages
library("tidyverse")
library("ggplot2")
library("dplyr")
library("knitr")
library(tseries) 
library(astsa)
```

```{r, include=FALSE}
#HELPER FUNCTIONS
get_scaled_metrics <- function(errors, train_series) {
  train_diffs <- abs(diff(as.numeric(train_series), lag = 1))
  nonseasonal_scale <- mean(train_diffs)
  nonseasonal_scale_sq <- mean(train_diffs^2)
  
  data.frame(
    MAE   = mean(abs(errors), na.rm = TRUE),
    RMSE  = sqrt(mean(errors^2, na.rm = TRUE)),
    MASE  = mean(abs(errors / nonseasonal_scale), na.rm = TRUE),
    MASSE = sqrt(mean(errors^2 / nonseasonal_scale_sq, na.rm = TRUE))
  )
}

get_arima_metrics <- function(model, x_train, x_test) {
  if (!is.null(model$pred)) {
    pred   <- as.numeric(model$pred)
    actual <- as.numeric(x_test)
    errors <- actual - pred
  } else {
    errors <- as.numeric(model$fit$residuals)
    errors <- errors[!is.na(errors)]
  }
  get_scaled_metrics(errors, x_train)
}

```

# 1 Introduction

Artificial Intelligence (AI) has been a hot topic in recent years, with it's wide applications in scientific discovery, advanced cybersecurity, and personalized AI agents. In the stock market, technology stock have accounted for 32% of global equity returns and 40% of total U.S. equity market returns since 2010 (Goldman Sachs, 2024), playing a dominant role in modern financial markets. Large-cap firms, such as Apple and Microsoft, represent a substantial share of total markets capitalization; consequently, their performance increasingly shapes the trajectory of major stock indices, including the S&P 500 and the NASDAQ Composite (Yahoo Finance). Since Technology companies are largely valued based on expectations of future innovation and scalability, breakthroughs can trigger significant revisions in investor sentiment, leading to movement in stock prices.

Breakthrough in new technology also impacts a firm's business model. For technology-centric companies in particular, launching a novel product pften increases company's net valuation (McKinsey, 2016). As several tech giants have launched flagship AI products, such as Gemini from Google and Claude from Anthropic, their market valuations and stock prices have shifted accordingly (Investing.com, 2026). The impact of novel AI product is seemly significant since it can impacts the company's valuation that the company does not even need to be publicly traded. The hard launch of ChatGPT in November 2022 significantly increased OpenAI's, a non-profit artificial intelligence research company, private valuation and attracted substantial investment, with its estimated worth skyrockeying from roughly \$20 billion to \$85.2 billion by 2026 (Barrientos, 2026).

Furthermore, the adoption of AI applications has also reshaped market narratives surrounding technological growth and productivity. Beyond the firms that directly launching AI models, companies perceived as key beneficiaries of AI infrastructure, most notably the GPU manufacturing company NVIDIA Corporation, have also experienced significant revaluations (Morningstar, 2026). This shift suggests that innovation shocks may influence not only short-term prices but also the broader structural dynamics of the overall technology sector.

With the crucial role of technology stock in the general financial market and AI product's impacts on company's evaluation, we want to investigate whether the technological breakthrough impact the technology sector stock market. This report investigates:

-   Did the stock return pattern changes after the emergence of AI products?

-   What's the predicted stock return based on current stock market trend?

Specifically, we evaluate this change by analyzing the Technology Select Sector SPDR Fund (XLK) using ARIMA models to examine whether the release of ChatGPT corresponds to a statistically significant shift in XLK price dynamics from 2017 to 2024.

Understanding whether such a technological breakthrough corresponds to a lasting change in market behavior is crucial for investors and policymakers. Persistent shifts in return pattern can affect risk assessment, portfolio allocation, and forecasting models in financial market analysis. Financial return series are time-dependent and often exhibit changes in volatility and growth dynamics. If the emergence of artificial intelligence has altered investor expectations, this effect may manifest as a structural break in the return-generating process.

# 2 Methodology

## 2.1 Data Overview

This report uses the daily XLK fund return price from Investing.com, a financial data platform that offers real-time data. The XLK stock price is collected by S&P Dow Jones Indices and State Street Global Advisory. S&P Dow Jones Indices is in charge of creating the company benchmark on entering the technology company stock pool. The State Street Global Advisors selected particular stocks into the combination of the XLK stock.

We chose the XLK because it's a comprehensive fund that captures an effective representation of the technology sector of the S&P 500 Index. The XLK has a 4-star in the Morningstar rating by March 2026, indicating an above-average performance that the XLK is in the top 22.5% to 32.5% of technology category. The XLK ETF provides precise exposure to companies from technology hardware, storage and peripherals; software; communications equipment; semiconductors and semiconductor equipment; IT services; and electronic equipment, instruments and components industries (State Street). This feature allows investors to take strategic or tactical positions at a more targeted level than traditional style based investing. The companies in the XLK are selected from the S&P 500 based on the Global Industry Classification Standard (GICS) where companies are assigned to a single classification based on revenue.

## 2.2 Data Collection

Using the dataset from the investing website, we derive the following variables for the analysis:

-   `Price` (in USD): The final price at which the ETF traded when the stock exchange closed (typically 4:00 PM EST).

-   `Open` (in USD): The price of the very first transaction when the market opened (9:30 AM EST).

-   `High` & `Low` (in USD): The maximum and minimum prices reached during the trading session.

-   `Vol.`: The total number of shares traded during the day (e.g., 15.90M means 15.9 million shares).

-   `Change`: The percentage difference between the current day's closing price and the previous day's closing price.

The dataset comprises daily observations spanning from January 3, 2017, to March 28, 2026. This time frame was specifically selected to ensure that the launch of ChatGPT (November 30, 2022) serves as a central pivot point, providing a balanced sample size for comparing pre- and post-innovation market dynamics

## 2.3 Data Cleaning

In the raw dataset from the investing website, we observe the point information on weekends are carried over from the Friday of the same week with empty number in `Vol`. We removed these points since they are not consistent with the general pattern and can be misleading in identifying the long-term price fluctuation. We then split the dataset into two groups, using the time stamp November 30, 2022 (the release date of chatGPT) as the cutoff.

```{r, include=FALSE}
#load data
data <- read.csv("XLK ETF Stock Price History.csv")
```

```{r, include=FALSE}
# change date 
data$Date <- as.Date(data$Date, format = "%m/%d/%Y") 
data <- data |>
  filter(trimws(Vol.) != "")

#split into two time frames
cutoff <- as.Date("2022-11-30")
before_data <- data[data$Date < cutoff, ]
after_data  <- data[data$Date >= cutoff, ]
```

```{r, include=FALSE}
#| eval: true
#| echo: false
#| message: false
#| warning: false
 # log return the raw data
data <- data |>
  mutate(log_price = log(Price),
         log_return = c(NA, diff(log_price)))

before_data <- data |> filter(Date < cutoff)
after_data  <- data |> filter(Date >= cutoff)
```

```{r, include=FALSE}
library(zoo)
library(lubridate) 
# [Link: Stack Overflow https://stackoverflow.com/questions/18755643/time-series-analysis-with-r-how-to-deal-with-daily-data]

# Create a zoo object
log_return_ts <- zoo(data$log_return[-1], order.by = data$Date[-1])

#remove NA
log_return_ts <- na.omit(log_return_ts)

# Plot the time series
#plot(log_return_ts, main = "Daily Time Series", xlab = "Date", ylab = "Log_return", type = "l", col = "blue")

# part 3.3 
x_train1 <- log_return_ts[index(log_return_ts) <= as.Date("2022-11-30")]
x_test1 <- log_return_ts[index(log_return_ts) > as.Date("2022-11-30")]

# part 3.4
#2321*0.2 # 464 2024-05-21
x_train2 <- log_return_ts[index(log_return_ts) <= as.Date("2024-05-21")]
x_test2 <- log_return_ts[index(log_return_ts) > as.Date("2024-05-21")]
```

```{r}
#| label: tbl-price
#| tbl-cap: "Summary Statistics of Stock Price" 
#| echo: false
#| message: false
#| warning: false
data |>
  summarise(
    mean_price = mean(Price, na.rm = TRUE),
    sd_price = sd(Price, na.rm = TRUE),
    min_price = min(Price, na.rm = TRUE),
    max_price = max(Price, na.rm = TRUE)
  ) |>
  kable(col.names = c("Mean Price", "Standard deviation", "Minimum", "Maximum"))
```

```{r}
#| label: fig-raw
#| fig-cap: "Raw Stock Price Over Time"
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(Date, Price)) +
  geom_line() +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "red") +
  labs(title = "Stock Price Over Time", subtitle = "Dashed line = 2022-11-30")
```

@tbl-price summarized the descriptive statistics for the stock price from January 3, 2017, to March 27, 2026, including the mean, standard deviation, and range. A more intuitive historical price trajectory is illustrated in @fig-raw, where the vertical red dashed line marks the official release of ChatGPT. Overall, the stock shows a sustained upward trend throughout the observed period.


```{r, include=FALSE}
#detrend
model_data <- data.frame(
  t = 1:nrow(data),
  y = data$Price
)

fit <- lm(y ~ t, data = model_data)
data$detrended_price <- resid(fit)
acf(data$detrended_price)
```

```{r}
#| label: fig-rawacf
#| fig-cap: "Autocorrelation Function (ACF) of Raw Stock Price."
#| eval: true
#| echo: false
#| message: false
#| warning: false
acf(data$Price, main="ACF of Raw Stock Price")
```


The autocorrelation function (ACF) plot in @fig-rawacf demonstrates a strong correlation between data points, indicating this time series is non-stationary and does not exhibits white noise characteristics. The ACF values persist well beyond the significance boundaries and decay at a very slow rate. This high correlation is explainable since the stock prices are accumulated based on the previous day's close price. Therefore, data transformation is required to achieve stationarity prior to modeling.



# 3 Methods and Results

## 3.1 Fitting a regression model on price
```{r, include=FALSE}
# #| eval: true
# #| echo: false
# #| message: false
# #| warning: false
#  # log return the raw data
# data <- data |>
#   mutate(log_price = log(Price),
#          log_return = c(NA, diff(log_price)))
# 
# before_data <- data |> filter(Date < cutoff)
# after_data  <- data |> filter(Date >= cutoff)
```

```{r}
#| label: tbl-log-return
#| tbl-cap: "Summary Statistics of Stock Price"
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| results: 'hide'
data |>
  mutate(period = ifelse(Date < cutoff, "Before AI", "After AI")) |>
  summarise(
    Mean = mean(log_return, na.rm = TRUE),
    SD = sd(log_return, na.rm = TRUE),
    .by = period
  ) |>
  mutate(
    Mean = round(Mean, 5),
    SD = round(SD, 5)
  ) |>
  kable()
```

We fit a regression model with a dummy variable $D_t$ as the predictor. 
$$\widehat{Price}=\hat\beta_0 + \hat\beta_1t + \hat\beta_2D_t + \hat\beta_3 tD_t$$ where $D_t=0$ represents the period before the appearance of ChatGPT, and $D_t=1$ represents the period after. 
<!-- The significance of $D_t$ and the interaction term $tD_t$ is used to assess whether there are differences between the two periods through a two sample mean test. The null hypothesis is that there is no difference in price behavior between the two periods, while the alternative hypothesis states that there is a difference.-->

```{r, include=FALSE}
data <- data[order(data$Date), ]
data$time_index <- 1:nrow(data)

data$AI_dummy <- ifelse(data$Date >= as.Date("2022-11-30"), 1, 0)
model <- lm(Price ~ time_index * AI_dummy, data = data)
```

```{r}
#| echo: false
#| results: 'hide'
mod_table <- summary(model)$coefficients
rownames(mod_table) <- c("Intercept", "time", "Dt", "time * Dt")
kable(
  round(mod_table, 3),
  caption = "Summary Table for Regression Model",
  col.names = c("Estimate", "Std. Error", "t value", "Pr(>|t|)")
) 
```

<!---Both the dummy variable and the interaction term are statistically significant, with p-values below 0.01, leading us to reject the null hypothesis. This indicates that prices behaved differently before and after the appearance of AI, not only in level but also in their trajectory over time.--->
As shown in @fig-price-reg-model, the two regression lines display different slopes across the two periods. The slope of price over time is steeper after the appearance of AI, meaning the price increased faster during this period.

```{r, fig.pos='H'}
#| label: fig-price-reg-model
#| fig-cap: "Fitted regression lines for price over time, separated by pre and post-AI periods."
#| eval: true
#| echo: false
#| message: false
#| warning: false
data$predicted <- predict(model, newdata = data)

ggplot(data, aes(Date, Price)) +
  geom_line() +
  geom_line(data = subset(data, AI_dummy == 0), 
            aes(y = predicted, color = "Before AI"), linewidth = 1) +
  geom_line(data = subset(data, AI_dummy == 1), 
            aes(y = predicted, color = "After AI"), linewidth = 1) +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("Before AI" = "steelblue", "After AI" = "tomato")) +
  labs(title = "Stock Price Over Time", subtitle = "Dashed line = 2022-11-30", color = "")
```

```{r, include=FALSE}
adf.test(model$residuals)
```

However, the assumptions on regression were not satisfied. The price data is non-stationary, and the QQ plot of the residuals indicates a clear deviation from normality. The  Augmented Dickey-Fuller (ADF) test on the model residuals yielded a p-value of 0.0575, failing to reject the null hypothesis of a unit root using 0.01 as the significance level. This suggests that the residuals remain non-stationary. 

A robust regression approach was considered too. However, since robust methods do not resolve issues of non-stationarity, this approach was not considered in further analysis. As the residuals are not stationary, an ARMA model cannot be appropriately fitted, and thus no further modelling was performed with regression.

```{r, include = FALSE}
# library(fixedCV)
# 
# robust <- robust_lm(model)
# robust$Summary_Table
```

## 3.2 Differencing on log price

To stabalize the variance and ensure stationarity, we applied a natural logarithmic transformation. The first difference of the log price series, known as the log return, was then used to remove the trend and produce a stationary series suitable for model estimation. @fig-logreturn shows the log return over time. While the log return mostly fluctuates around a mean of zero with small fluctuations, there are few large spikes. The biggest fluctuations occur around 2020 when the pandemic began, suggesting that the pandemic can might casue substantial uncertainty and sharp movements in the market.

```{r}
#| label: fig-logreturn
#| fig-cap: "Daily Log Returns of Stock Price (2017–2026)."
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(x = Date, y = log_return)) +
  geom_line(linewidth = 0.4) +
  geom_vline(xintercept = cutoff, linetype = "dashed", color ="red") +
  labs(title = "Daily Log Returns of Stock Price",
    subtitle = "Vertical dashed line indicates ChatGPT release on November 30, 2022",
    x = "Year", y = "Log Return")
```

```{r}
#| label: fig-logacfpacf
#| fig-cap: "Autocorrelation Function (ACF) and Partial Autocorrelation Function (PACF) of Log Return."
#| eval: true
#| echo: false
#| message: false
#| warning: false

par(mfrow = c(1,2))
acf(na.omit(data$log_return), main="ACF of log return")
pacf(na.omit(data$log_return), main="PACF of log return")
```

```{r, include=FALSE}
adf.test(na.omit(data$log_return))
```
Although, in @fig-logacfpacf, the ACF shows there still exists few autocorrelation coefficients outside the significance boundaries, the ADF test was conducted for a formal assessment. The test yielded a p-value of 0.01, providing strong empirical evidence to reject the null hypothesis of a unit root and confirming that the log return series is stationary.

## 3.3 Using Pre-AI Prices to Predict Post-AI Prices
 
We aimed to examine whether the emergence of AI had an unexpected impact on the stock market and contributed to increased price fluctuations. To investigate this, we used stock price data from January 3, 2017 to November 30, 2022 to forecast stock prices after November 30, 2022.

First, we fitted ARIMA models using the historical price data from January 3, 2017 to November 30, 2022. We used AIC, AICc and BIC metrics to determine the best model for forecasting. 

We then applied three "benchmark" forecasting methods: the mean method, the naive method, and the seasonal naive method. Mean method forecasts all future values as equal to the average of the training set. Naive method set all forecast values to be the value of the last observation of the historical data. Seasonal naive method set each forecast to be equal to the last observed value from the same season. In this case, we tested for weekly seasonality using this method. 

Finally, we compared the performance of the ARIMA model and the benchmark methods using point comparison metrics like MAE, RMSE, MASE and MASSE.


```{r, include=FALSE}
part <- ts(as.numeric(log(before_data$Price)))

acf(diff(log(before_data$Price)))
pacf(diff(log(before_data$Price)))
```


Using the ACF and PACF plots, we can determine the appropriate order of the ARIMA model. According to @fig-logacfpacf, the ACF gradually tails off, suggesting the presence of autoregressive behavior. According to the PACF, there are strong spikes at lags 2, 4, 7, and 8, and several other significant lags. Based on these patterns, ARIMA(1,1,1), ARIMA(2,1,2), ARIMA(4,1,4), ARIMA(6,1,6), ARIMA(7,1,7), and ARIMA(8,1,8) models were fitted to the log price data to capture the time series.

The p values for Ljung-Box Statistic plots indicate that ARIMA(1,1,1) and ARIMA(2,1,2) are not good fits, because p values are significant for all lags. ACF plots for ARIMA (4,1,4), ARIMA(7,1,7), and ARIMA (8,1,8) looks stationary, but QQ plots show the points are deviated from the reference line, meaning the normality condition is not met. To determine which model performs the best, we compare the model AIC, AICc, BIC, and log likelihood values.

```{r, include=FALSE}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| results: 'hide'
model11 <- sarima(part, p=1, d=1, q=1, no.constant=TRUE, details = FALSE)
model22 <- sarima(part, p=2, d=1, q=2, no.constant=TRUE, details = FALSE)
model33 <- sarima(part, p=4, d=1, q=4, no.constant=TRUE, details = FALSE)
model44 <- sarima(part, p=6, d=1, q=6, no.constant=TRUE, details = FALSE)
model55 <- sarima(part, p=7, d=1, q=7, no.constant=TRUE, details = FALSE)
model66 <- sarima(part, p=8, d=1, q=8, no.constant=TRUE, details = FALSE)
```

```{r}
#| label: tbl-arima-comparison
#| tbl-cap: "Comparison of Model Performance"
#| eval: true
#| echo: false
#| message: false
#| warning: false
part_ICs <- rbind(
  model11$ICs,
  model22$ICs,
  model33$ICs,
  model44$ICs,
  model55$ICs,
  model66$ICs
)

part_loglik <- c(
  logLik(model11$fit),
  logLik(model22$fit),
  logLik(model33$fit),
  logLik(model44$fit),
  logLik(model55$fit),
  logLik(model66$fit)
)
part_results <- cbind(
  part_ICs,
  "Log Likelihood" = part_loglik
)
rownames(part_results) <- c("ARIMA(1,1,1)", "ARIMA(2,1,2)","ARIMA(4,1,4)", "ARIMA(6,1,6)", "ARIMA(7,1,7)", "ARIMA(8,1,8)")

part_results |> kable()
```

```{r, include= FALSE}
cat("train length:", length(x_train1), "\n")
cat("test length:", length(x_test1), "\n")

x_train_part <- ts(as.numeric(x_train1))
x_test_part  <- ts(as.numeric(x_test1))

fit111_part <- sarima.for(x_train_part, n.ahead=length(x_test_part), p=1, d=1, q=1, no.constant=TRUE)
fit212_part <- sarima.for(x_train_part, n.ahead=length(x_test_part), p=2, d=1, q=2, no.constant=TRUE)
fit414_part <- sarima.for(x_train_part, n.ahead=length(x_test_part), p=4, d=1, q=4, no.constant=TRUE)
fit616_part <- sarima.for(x_train_part, n.ahead=length(x_test_part), p=6, d=1, q=6, no.constant=TRUE)
fit717_part <- sarima.for(x_train_part, n.ahead=length(x_test_part), p=7, d=1, q=7, no.constant=TRUE)
fit818_part <- sarima.for(x_train_part, n.ahead=length(x_test_part), p=8, d=1, q=8, no.constant=TRUE)

forecast_models <- list(
  "ARIMA(1,1,1)" = fit111_part,
  "ARIMA(2,1,2)" = fit212_part,
  "ARIMA(4,1,4)" = fit414_part,
  "ARIMA(6,1,6)" = fit616_part,
  "ARIMA(7,1,7)" = fit717_part,
  "ARIMA(8,1,8)" = fit818_part
)

part_results <- do.call(rbind, lapply(forecast_models, get_arima_metrics, x_train = x_train_part, x_test  = x_test_part))

```

We also compared the MAE, RMSE, MASE, and MASSE for all ARIMA models. There was no significant difference between models. Based on the AICc values, ARIMA (7,1,7) has the lower AICc values. Therefore, ARIMA(7,1,7) was picked as the best ARIMA model for forecasting. 

```{r, include=FALSE}
#| results: 'hide'
part_results |> kable(digits=5)
```


```{r}
#| eval: true
#| echo: false
#| fig-cap: "Diagnostic Plot for ARIMA(7,1,7)"
#| message: false
#| warning: false
#| results: 'hide'
sarima(part, p=7, d=1, q=7, no.constant=TRUE)
```





```{r, include=FALSE}
#adf.test(log_return_ts)
```

```{r, include=FALSE}
#| eval: true
#| echo: false
#| message: false
#| warning: false
x_train1 <- log_return_ts[index(log_return_ts) <= as.Date("2022-11-30")]
x_test1  <- log_return_ts[index(log_return_ts) > as.Date("2022-11-30")]

#forecasts
x_mean1 <- rep(mean(x_train1), length(x_test1)) #mean
x_naive1 <- rep(tail(x_train1,1), length(x_test1)) #naive
x_seasonal1 <- rep(c(tail(x_train1, 7)), length(x_test1)) #seasonal

bench_part <- rbind(
  Mean     = get_scaled_metrics(as.numeric(x_test1) - x_mean1, x_train1),
  Naive    = get_scaled_metrics(as.numeric(x_test1) - x_naive1, x_train1),
  Seasonal = get_scaled_metrics(as.numeric(x_test1) - x_seasonal1, x_train1)
)

bench_part |> kable(digits=5, caption="Benchmark Forecast Errors-before ai")

#forecast errors
mean_errors1 <- x_test1 - x_mean1
naive_errors1 <- x_test1 - x_naive1
seasonal_errors1 <- x_test1 - x_seasonal1

# MAE(Scale-dependent Errors)
mean_MAE1 <- mean(abs(mean_errors1), na.rm = TRUE)
naive_MAE1 <- mean(abs(naive_errors1), na.rm = TRUE)
seasonal_MAE1 <- mean(abs(seasonal_errors1), na.rm = TRUE)

#RMSE
mean_RMSE1 <- sqrt(mean(mean_errors1^2, na.rm = TRUE))
naive_RMSE1 <- sqrt(mean(naive_errors1^2, na.rm = TRUE))
seasonal_RMSE1 <- sqrt(mean(seasonal_errors1^2, na.rm = TRUE))
```

```{r, include=FALSE}
cat("MAE - Mean Forecast:     ", mean_MAE1, "\n")
cat("MAE - Naive Forecast:    ", naive_MAE1, "\n")
cat("MAE - Seasonal Forecast: ", seasonal_MAE1, "\n\n")

cat("RMSE - Mean Forecast:     ", mean_RMSE1, "\n")
cat("RMSE - Naive Forecast:    ", naive_RMSE1, "\n")
cat("RMSE - Seasonal Forecast: ", seasonal_RMSE1, "\n")
```

```{r, include=FALSE}
n_train_part <- length(x_train1)
train_diffs_part  <- abs(diff(as.numeric(x_train1), lag = 1))
nonseasonal_scale_part <- mean(train_diffs_part)
nonseasonal_scale_sq_part <- mean(train_diffs_part^2)

# for(i in 2:n){
#   train_MAE <- abs(x_train2[i] - x_train2[i-1])
# }
# nonseasonal_scale <- sum(train_MAE)/(n-1)
# nonseasonal_scale_sq <- sum(train_MAE^2)/(n-1)

mean_MASE1 <- mean(abs(mean_errors1/nonseasonal_scale_part), na.rm = TRUE)
naive_MASE1 <- mean(abs(naive_errors1/nonseasonal_scale_part), na.rm = TRUE)
seasonal_MASE1 <- mean(abs(seasonal_errors1/nonseasonal_scale_part), na.rm = TRUE)

mean_MASSE1 <- sqrt(mean(mean_errors1^2/nonseasonal_scale_sq_part, na.rm = TRUE))
naive_MASSE1 <- sqrt(mean(naive_errors1^2/nonseasonal_scale_sq_part, na.rm = TRUE))
seasonal_MASSE1 <- sqrt(mean(seasonal_errors1^2/nonseasonal_scale_sq_part, na.rm = TRUE))
```

```{r}
#| label: tbl-forecast_errors
#| eval: true
#| echo: false
#| message: false
#| warning: false

arima717_part <- part_results[rownames(part_results) == "ARIMA(7,1,7)", ]

forecast_table_part <- rbind(bench_part, arima717_part)
kable(forecast_table_part, digits=5, caption="Forecast Errors")
```

According to @tbl-forecast_errors, MAE (Mean Absolute Error) and RMSE (Root Mean Squared Error), which measure the difference between actual and predicted values, are small for the mean method but also for naive and ARIMA models. Hence, a simple model like Mean method can effectively predict the data following the cutoff date using data prior to November 30, 2022. We assumed ARIMA would do a better job, because it's a more comprehensive model. But our results showed that simple model can be powerful and was good enough for this prediction.

Furthermore, this suggests that AI have not yet contribute a significant impact in the stock market, at least in terms of predictability using our proposed models. As a result, we will use the entire time frame, allocating 80% for training and 20% for testing, to build a forecast model for predicting prices in 2026.


# 3.4 Comparing Models Over the Full Time Frame
80% of data from January 3, 2017 to March 27, 2026 were used as the training set, and 20% of data were used for testing in building our models. Similar to section 3.3, ARIMA(1,1,1), ARIMA(2,1,2), ARIMA(4,1,4), ARIMA(7,1,7), and ARIMA(8,1,8) models were fitted. The best ARIMA model was determined using AIC, AICc, BIC, and log liklihood. In addition, three benchmark methods -- the mean method, naive method, and seasonal naive method -- were applied for comparison. The best-performing ARIMA model was then compared against these benchmark methods.

First, we fit the proposed ARIMA models. The diagnostics plots suggested that ARIMA(4,1,4), ARIMA(7,1,7) and ARIMA (8,1,8) might be the better model, because some p values of the Ljung box test were insignificant. To formally determine the most appropriate model structure, we compared the candidate models using the criteria shown in @tbl-model-comparison. 
```{r, include=FALSE}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| results: 'hide'
model1 <- sarima(data$log_price, p=1, d=1, q=1, no.constant=TRUE) 
model2 <- sarima(data$log_price, p=2, d=1, q=2, no.constant=TRUE) 
model3 <- sarima(data$log_price, p=4, d=1, q=4, no.constant=TRUE)
model4 <- sarima(data$log_price, p=7, d=1, q=7, no.constant=TRUE)
model5 <- sarima(data$log_price, p=8, d=1, q=8, no.constant=TRUE)
```

```{r}
#| label: tbl-model-comparison
#| tbl-cap: "Comparison of Model Performance"
#| eval: true
#| echo: false
#| message: false
#| warning: false
all_ICs <- rbind(
  model1$ICs,
  model2$ICs,
  model3$ICs,
  model4$ICs,
  model5$ICs
)

all_loglik <- c(logLik(model1$fit), logLik(model2$fit), 
                logLik(model3$fit), logLik(model4$fit), 
                logLik(model5$fit))

all_results <- cbind(
  all_ICs,
  logLik = all_loglik
)
rownames(all_results) <- c("ARIMA(1,1,1)", "ARIMA(2,1,2)","ARIMA(4,1,4)", "ARIMA(7,1,7)", "ARIMA(8,1,8)")

all_results |> kable()
```

```{r, include=FALSE}
x_train_ts <- ts(as.numeric(x_train2))
x_test_ts  <- ts(as.numeric(x_test2))

fit111_full <- sarima.for(x_train_ts, n.ahead=length(x_test_ts), p=1, d=1, q=1, no.constant=TRUE)
fit212_full <- sarima.for(x_train_ts, n.ahead=length(x_test_ts), p=2, d=1, q=2, no.constant=TRUE)
fit414_full <- sarima.for(x_train_ts, n.ahead=length(x_test_ts), p=4, d=1, q=4, no.constant=TRUE)
fit717_full <- sarima.for(x_train_ts, n.ahead=length(x_test_ts), p=7, d=1, q=7, no.constant=TRUE)
fit818_full <- sarima.for(x_train_ts, n.ahead=length(x_test_ts), p=8, d=1, q=8, no.constant=TRUE)

forecast_models_full <- list(
  "ARIMA(1,1,1)" = fit111_full,
  "ARIMA(2,1,2)" = fit212_full,
  "ARIMA(4,1,4)" = fit414_full,
  "ARIMA(7,1,7)" = fit717_full,
  "ARIMA(8,1,8)" = fit818_full
)

full_results <- do.call(rbind, lapply(forecast_models_full, get_arima_metrics, x_train = x_train_ts, x_test  = x_test_ts))
```

```{r, include=FALSE}
full_results |> kable(digits=5)
```

Among the candidate models, ARIMA(4,1,4) was selected as the best fitting model, as it achieved the lowest AIC, AICc, and BIC values. While ARIMA(7,1,7) produced a very similar AIC, its notably higher BIC indicates that the additional parameters are not justified, making ARIMA(4,1,4) the preferred specification.

```{r}
#| eval: true
#| echo: false
#| fig-cap: "Diagnostic Plot for ARIMA(4,1,4)"
#| message: false
#| warning: false
#| results: 'hide'
model3 <- sarima(data$log_price, p=4, d=1, q=4, no.constant=TRUE)
```



```{r, include=FALSE}
# entire data set 2017/1/3 - 2026/3/27

# #2321*0.2 # 464 2024-05-21
#
x_train2 <- log_return_ts[index(log_return_ts) <= as.Date("2024-05-21")]
x_test2 <- log_return_ts[index(log_return_ts) > as.Date("2024-05-21")]

#forecasts
x_mean2 <- rep(mean(x_train2), length(x_test2)) #mean
x_naive2 <- rep(tail(x_train2,1), length(x_test2)) #naive
x_seasonal2 <- rep(c(tail(x_train2, 7)), length(x_test2)) #seasonal

# # forecast errors
# mean_errors2 <- x_test2 - x_mean2
# naive_errors2 <- x_test2 - x_naive2
# seasonal_errors2 <- x_test2 - x_seasonal2
# 
# # MAE(Scale-dependent Errors)
# mean_MAE2 <- mean(abs(mean_errors2), na.rm = TRUE)
# naive_MAE2 <- mean(abs(naive_errors2), na.rm = TRUE)
# seasonal_MAE2 <- mean(abs(seasonal_errors2), na.rm = TRUE)
# 
# #RMSE
# mean_RMSE2 <- sqrt(mean(mean_errors2^2, na.rm = TRUE))
# naive_RMSE2 <- sqrt(mean(naive_errors2^2, na.rm = TRUE))
# seasonal_RMSE2 <- sqrt(mean(seasonal_errors2^2, na.rm = TRUE))

bench_full <- rbind(
  Mean     = get_scaled_metrics(x_test2 - x_mean2,     x_train2),
  Naive    = get_scaled_metrics(x_test2 - x_naive2,    x_train2),
  Seasonal = get_scaled_metrics(x_test2 - x_seasonal2, x_train2)
)

# bench_full |> kable(digits=5, caption="Benchmark Forecast Errors - full")
```

```{r, include=FALSE}
# # problem here with MAPE b/c our data is around 0, so it blew up
# valid <- x_test2 != 0 & !is.na(x_test2)
# 
# # percentage error
# mean_p_error <- 100*mean_errors2/x_test2[valid]
# naive_p_error <- 100*naive_errors2/x_test2[valid]
# seasonal_p_error <- 100*seasonal_errors2/x_test2[valid]
# 
# #mean absolute percentage error
# mean_MAPE <- mean(abs(mean_p_error), na.rm = TRUE)
# naive_MAPE <- mean(abs(naive_p_error), na.rm = TRUE)
# seasonal_MAPE <- mean(abs(seasonal_p_error), na.rm = TRUE)
```

```{r, include=FALSE}
# n_train <- length(x_train2)
# train_diffs <- abs(diff(as.numeric(x_train2), lag = 1))
# nonseasonal_scale <- mean(train_diffs)
# nonseasonal_scale_sq <- mean(train_diffs^2)
# 
# # for(i in 2:n){
# #   train_MAE <- abs(x_train2[i] - x_train2[i-1])
# # }
# # nonseasonal_scale <- sum(train_MAE)/(n-1)
# # nonseasonal_scale_sq <- sum(train_MAE^2)/(n-1)
# 
# mean_MASE <- mean(abs(mean_errors2/nonseasonal_scale), na.rm = TRUE)
# naive_MASE <- mean(abs(naive_errors2/nonseasonal_scale), na.rm = TRUE)
# seasonal_MASE <- mean(abs(seasonal_errors2/nonseasonal_scale), na.rm = TRUE)
# 
# mean_MASSE <- sqrt(mean(mean_errors2^2/nonseasonal_scale_sq, na.rm = TRUE))
# naive_MASSE <- sqrt(mean(naive_errors2^2/nonseasonal_scale_sq, na.rm = TRUE))
# seasonal_MASSE <- sqrt(mean(seasonal_errors2^2/nonseasonal_scale_sq, na.rm = TRUE))
```

```{r}
#| label: tbl-forecast_errors_full
#| eval: true
#| echo: false
#| message: false
#| warning: false

# forecast_table <- data.frame(
#   Method = c("Mean", "Naive", "Seasonal", "ARIMA(4,1,4)"),
#   MAE = c(mean_MAE2, naive_MAE2, seasonal_MAE2, arimaMAE),
#   RMSE = c(mean_RMSE2, naive_RMSE2, seasonal_RMSE2, arimaRMSE),
#   MASE = c(mean_MASE, naive_MASE, seasonal_MASE, arimaMASE), #xs was meanMASE
#   MASSE = c(mean_MASSE, naive_MASSE, seasonal_MASSE, arimaMASSE)
# )
#
# kable(forecast_table, digits = 5, caption = "Forecast Errors-full model")
arima414_full <- full_results[rownames(full_results) == "ARIMA(4,1,4)", ]

forecast_table_full <- rbind(bench_full, arima414_full)
kable(forecast_table_full, digits=5, caption="Forecast Errors - Full Model")
```

@tbl-forecast_errors_full shows the naive method has the lowest MAE while the mean method has the lowest RMSE. MAPE (Mean Absolute Percentage Error) is a percentage-based error metric that scales forecast errors relative to the actual values. However, MAPE(Mean Absolute Percentage Error) is not included in this comparison because the log return data fluctuates around zero, where some small values closer to zero make MAPE enormous and not a good parameter to look at. MASE (Mean Absolute Scaled Error) and MASSE (Root Mean Squared Scaled Error) are scaled error metrics, and they works by scaling errors based on the training MAE from a simple forecast method. This makes them stable because the denominator is a fixed constant derived from the training data, providing a more informative alternative when MAPE fails to give meaningful results. Based on the lowest MASE and RMSSE values, the mean method and ARIMA (4,1,4) performs best overall. Again, the simple models performed as good as ARIMA. The differences between models were extremely small. ARIMA model may not be needed to do forecasting in this case.



```{r}
#| label: fig-qqhist
#| fig-cap: "QQ plot and Histogram for log returns"
#| eval: true
#| echo: false
#| message: false
#| warning: false
par(mfrow = c(1,2))
qqnorm(log_return_ts, main = "QQ plot")

hist(log_return_ts, main = "histogram")
```

Next, we construct a prediction interval for the mean method forecast model and ARIMA(4,1,4). In @fig-qqhist, while the histogram suggests the residuals are approximately normally distributed, the QQ-plot indicates that the log returns deviate from normality, violating classic assumptions. Despite this, @fig-full-mean-prediction-int shows that the prediction interval successfully captures all observations in the test set, suggesting the interval is functioning as intended. However, the interval is extremely wide, reflecting high uncertainty in the forecasts. This indicates that while the mean method provides valid coverage, it offers little practical predictive power for long-term returns.

```{r}
#| label: fig-full-mean-prediction-int
#| fig-cap: "Mean method forecasts and prediction intervals for log returns"
#| eval: true
#| echo: false
#| message: false
#| warning: false
#prediction interval
forecast_time <- time(x_test2)
forecast <- x_mean2 

alpha <- 0.05
sigma <- sqrt(sum((log_return_ts - mean(log_return_ts))^2) / 
              (length(log_return_ts) - 1))
c <- qnorm(1-alpha/2)
sigma_h <- sigma*sqrt(1:length(forecast_time))
LB <- forecast - c*sigma_h
UB <- forecast + c*sigma_h

original_data <- data.frame(points = c(log_return_ts), time = c(time(log_return_ts)))

forecast_data <- data.frame(points = forecast, time = forecast_time, LB = LB, UB = UB)

ggplot(data = original_data, aes(x = time, y = points)) + geom_line( ) +
  ylab("Log Returns") + 
  geom_line(data = forecast_data, aes(x = time, y = points), colour = "blue") + 
  geom_ribbon(data = forecast_data, aes(x = time, ymin = LB, ymax = UB), fill = "lightblue", alpha = 0.5)

```

@fig-arima-forecast shows the forecast results using ARIMA (4,1,4) compared with the actual data. The forecast of ARIMA was close to zero. This was because the log returns had a mean close to zero. However, the prediction interval did not capture all the actual data, especially during period of extreme volatility. Therefore, while ARIMA is effective in predicting the average behavior, it struggles to predict sudden large change in the data. 

```{r}
#| label: fig-arima-forecast
#| fig-cap: "ARIMA(4,1,4) Forecast Data Compared with Actual Data"
#| eval: true
#| echo: false
#| message: false
#| warning: false
library(forecast)

# convert keeping dates
fit414_Arima <- Arima(as.numeric(x_train2), order=c(4,1,4), include.mean=FALSE)
fc414 <- forecast(fit414_Arima, h=length(x_test2))

# build data frames with actual dates
df_train <- data.frame(date=index(x_train2), value=as.numeric(x_train2))
df_test  <- data.frame(date=index(x_test2),  value=as.numeric(x_test2))
df_fc    <- data.frame(
  date  = index(x_test2),
  mean  = as.numeric(fc414$mean),
  lo95  = as.numeric(fc414$lower[,2]),
  hi95  = as.numeric(fc414$upper[,2])
)

ggplot() +
  geom_line(data=df_train, aes(x=date, y=value), color="black") +
  geom_line(data=df_test, aes(x=date, y=value, color="Actual")) +
  geom_line(data=df_fc, aes(x=date, y=mean,  color="Forecast")) +
  geom_ribbon(data=df_fc,  aes(x=date, ymin=lo95, ymax=hi95), fill="lightblue", alpha=0.5) +
  scale_color_manual(values=c("Actual"="red", "Forecast"="blue")) +
  ggtitle("ARIMA(4,1,4) Forecast vs Test Set") +
  labs(title="ARIMA(4,1,4) Forecast vs Test Set", x="Date", y="Log Return", color="") + theme(legend.position=c(1,1), legend.justification=c(1,1))
```



## 3.5 Forecast Future 2026 Data
Using ARIMA(4,1,4), we forecast the next 30 days starting March 28, 2026. The predicted values appeared relatively flat, which is expected for ARIMA models over short period of time. Since future white noise terms are unknown, the forecasts tend to settle toward a constant, resulting in a smoother and less stochastic pattern compared to the real data, as shown in @fig-full-arima-forecast. Although this is a simple model, the forecasts are still useful for understanding the potential short-term behavior and overall trend of the market. Investors and analysts may use these forecasts as a reference for anticipating future movement patterns, and supporting short-term decision making.

```{r, include=FALSE}
forecasted <- sarima.for(ts(data$log_price), n.ahead = 30, p = 4, d = 1, q = 4, no.constant = TRUE)$pred
final_forecasted <- exp(forecasted)
```

```{r}
#| label: fig-full-arima-forecast
#| fig-cap: "ARIMA(4,1,4) fitted values and forecasts from 2017 to 2026, with a closer view of the 2026 forecast period"
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 6

#https://www.geeksforgeeks.org/r-language/sarima-in-r/

par(mfrow = c(2, 1))

forecast_dates <- seq(max(data$Date) + 1, by = "day", length.out = 30)
forecast_df <- data.frame(Date = forecast_dates, Price = final_forecasted)

# Full period
plot(data$Date, data$Price, type = "l",
     xlab = "Date", ylab = "Price", main = "Price Forecast")
points(forecast_df$Date, forecast_df$Price, col = "red")

# 2026 
xlim_range <- c(as.Date("2026-01-01"), max(forecast_dates))
plot(data$Date, data$Price, type = "l",
     xlab = "Date", ylab = "Price", main = "2026 Forecast",
     xlim = xlim_range)
points(forecast_df$Date, forecast_df$Price, col = "red")
```

\newpage

# 4 Conclusions

This project used stock price data from January 3, 2017 to March 27, 2026 to study stock price behavior and forecasting performance using ARIMA models and benchmark forecasting methods. The first part of the project investigated whether the emergence of AI, specifically following the launch of ChatGPT on November 30, 2022, had a meaningful impact on stock market. To examine this question, we used data before November 30, 2022 to forecast stock prices after this date. Several ARIMA models and three benchmark forecasting methods were proposed and evaluated. The results showed that the performance of the best performing ARIMA models was similar to that of the simpler benchmark methods. Overall, the models were able to forecast the post-AI period reasonably well using pre-AI data, suggesting that the introduction of AI did not produce a significant change in the stock price. Therefore, we used the full data frame for forecasting. 

In the second part, we used the full time frame from January 3, 2017 to March 27, 2026 to further evaluate forecasting performance. Multiple ARIMA models and benchmark methods were compared accuracy metrics. The results again indicated no substantial difference in predictive performance among the models. Among the ARIMA models, ARIMA(4,1,4) was selected as the preferred model based on model selection criteria, and it was used to forecast stock prices for the following 30 days.



# 5 Limitations

There are some potential limitations for the dataset. One of the limitations is that there's possibility for State Street to rebalance the stock composition over time, which means the dataset may not remain fully consistent throughout the study period. The companies in the XLK ETF are selected based on the GICS. However, the GICS structure is reviewed annually, meaning State Street need to adjust the XLK portfolio at least on a yearly base. This could reduce the reliability of comparing stock performance across different years and weaken the robustness of analysis.

Another limitation is the dependency of the dataset. In the data analysis, we noticed few large spikes in the log return plot, suggeting factors like pandemics may also attributes to the stock price fluctuation. In Section 3, we found out the time series data points remain highly correlated even after the data transformation. For instance, the QQ plot indicates deviations from normality, suggesting that model assumptions may not hold and prediction intervals may be unreliable. In addition, our model focuses on the short-term impacts after the release of ChatGPT, but the effect of AI may be better reflected in a long term trend rather than a boom. The log transformation may mask the underlying long term trend, because it removes the trend to make the data stationary. The log transformation only focuses on daily changes rather than long-term structural changes. As a results, model based on the log transformation is better at studying short-term market reaction but may be less effective at capturing long-term changes in the market over time.

Also, AR, MA, ARMA are all models that are only a function of the time series itself. They do not include other important factors, such as government policy or economic policy changes. There are many other factors that affects the stock market that we did not incorporate, such as economic indicators and interest rates. Since stock market is influenced by many external variables, excluding these factors may limit the strength and forecasting accuracy of the models.

# 6 Future Directions

To account for external factors that might affect the stock market, we could add these as predictors in our model. This can help us better estimate stock prices.

Although we have not been able to conclude AI has significant impact on the behavior of technology sector stock returns, it is still useful to estimate how sensitive technology sector stock price comparing to the broader market indices or other sectors. We might expect the technology sector to show higher volatility and respond more to growth expectations. In future studies, we can fit and compare separate models for different market segments, such as the S&P 500 and individual sectors like healthcare or financials, to better understand these differences.

It is also possible to incorporate the method proposed by Campos-Martins (2024) with our model. After identifying notable development by AI companies, we can adjust our predictors to improve accuracy. For example, we could include an indicator variable that captures major AI events or achievements. This would allow the model to account for changes over time driven by new innovations, instead of relying only on fixed predictors.

Finally, our research should follow a consistent process where we constantly refit the models as the stock market changes over time. Updating the model with new data can help us see whether there are any long-term effects or clear shifts in the data, especially as AI continues to develop and become more widely used.
