---
title: "main"
format: pdf
---
```{r, include=FALSE}
#packages
library("tidyverse")
library("ggplot2")
library("dplyr")
```

# Section 1. Data Collection
The data are from 1/28/2019 to 3/28/2026. 


# Section 2. Variable Creation and Data Cleaning
Original dataset contains date as a character which may cause problems when using it as our time predictor, so we converted to the Date type in R. 
We observed outliers that are doubled the price of the remaining data points. Looking at the data, we found that for most weekends, prices are carried over from the Friday of the same week. These outliers correspond to weekend days, as indicated by their dates, and their volumes are recorded as empty strings. We decided to remove these outliers because they are not consistent with the general pattern, which they doubled the Friday price. 
Lastly, we split the dataset into two groups using November 30, 2022 (the release date of ChatGPT) as the cutoff. This is used for the comparison of the two time seires.

We will detrend the data (and difference) the data to remove the overall growth in the stock market and possible market cycles. 
```{r, include=FALSE}
#load data
data <- read.csv("XLK ETF Stock Price History.csv")
```

```{r, include=FALSE}
# change date 
data$Date <- as.Date(data$Date, format = "%m/%d/%Y") 
data <- data |>
  filter(trimws(Vol.) != "")
```

```{r, include=FALSE}
#split into two time frames
cutoff <- as.Date("2022-11-30")
before_data <- data[data$Date < cutoff, ]
after_data  <- data[data$Date >= cutoff, ]
```

# Section 3 Analytic Methods 
## regression model
### model assumption check

## detrend 
- check if data is white noise, or MA, or AR
```{r}
plot(data$Date, data$Price, type = "l")
```

```{r}
par(mfrow = c(2,2))
plot(before_data$Date, before_data$Price, type = "l")
plot(after_data$Date, after_data$Price, type = "l", col = "blue")
```
```{r}
#detrend
model_data <- data.frame(
  t = 1:nrow(data),
  y = data$Price
)

fit <- lm(y ~ t, data = model_data)

```

## forecasting 

# Section 4

