---
title: "main"
format: pdf
---
```{r, include=FALSE}
#packages
library("tidyverse")
library("ggplot2")
library("dplyr")
library("knitr")
```

# Section 1. Data Collection
The XLK stock price we are looking at is accounted by S&P Dow Jones Indices and State Street Global Advisory. S&P Dow Jones Indices is in charge of creating the company benchmark on entering the technology company stock pool; while the State Street Global Advisors selected particular stocks into the combination of the XLK stock. The data we are using is collected from the investing.com website, a financial data platform that provides public real-time market data directly from stock exchanges such as the NYSE and NASDAQ. The primary purpose of collecting the data is to provide a transparent record of the performance of the Technology Selected Sector Index. The subjects of this data series are the large-cap U.S. technology companies such as  Microsoft, NVIDIA, and Apple whose market valuations and returns constitute the movement of the XLK fund.

The variables we are using includes: 

- Date: The specific trading day of the observed data

- Price (Close): The final price at which the ETF traded when the stock exchange closed (typically 4:00 PM EST).

- Open:  The price of the very first transaction when the market opened (9:30 AM EST).

- High & Low: The maximum and minimum prices reached during the trading session.

- Vol.: The total number of shares traded during the day (e.g., 15.90M means 15.9 million shares).

- Change: The percentage difference between the current day's closing price and the previous day's closing price.

The data we are using is collected from 1/28/2019 to 3/28/2026, which incorporates the AI outbreak time. One of the limitations of this dataset is that there’s a possibility for State Street to rebalance the stock composition over time, thus the data might not be consistent to offer a robustic overtime analysis. 


# Section 2. Variable Creation and Data Cleaning
Original dataset contains date as a character which may cause problems when using it as our time predictor, so we converted to the Date type in R. 
We observed outliers that are doubled the price of the remaining data points. Looking at the data, we found that for most weekends, prices are carried over from the Friday of the same week. These outliers correspond to weekend days, as indicated by their dates, and their volumes are recorded as empty strings. We decided to remove these outliers because they are not consistent with the general pattern, which they doubled the Friday price. 
Lastly, we split the dataset into two groups using November 30, 2022 (the release date of ChatGPT) as the cutoff. This is used for the comparison of the two time seires.

We will difference the data to stabilize variance, and remove the overall growth in the stock market as well as possible market cycles. 
```{r, include=FALSE}
#load data
data <- read.csv("XLK ETF Stock Price History.csv")
```

```{r, include=FALSE}
# change date 
data$Date <- as.Date(data$Date, format = "%m/%d/%Y") 
data <- data |>
  filter(trimws(Vol.) != "")
```

```{r, include=FALSE}
#split into two time frames
cutoff <- as.Date("2022-11-30")
before_data <- data[data$Date < cutoff, ]
after_data  <- data[data$Date >= cutoff, ]
```

# Section 3 Analytic Methods 
### model assumption check
- check if data is white noise, or MA, or AR
```{r, include=FALSE}
plot(data$Date, data$Price, type = "l", xlab="Date", ylab="Price", main = "Stock price over time (2019 - 2026)")
```

```{r, include=FALSE}
par(mfrow = c(1,2))
plot(before_data$Date, before_data$Price, type = "l")
plot(after_data$Date, after_data$Price, type = "l", col = "blue")
```

```{r, include=FALSE}
#detrend
model_data <- data.frame(
  t = 1:nrow(data),
  y = data$Price
)

fit <- lm(y ~ t, data = model_data)
data$detrended_price <- resid(fit)
acf(data$detrended_price)
```

# Section 4
## EDA
```{r}
#| label: tbl-price
#| tbl-cap: "Summary Statistics of Stock Price"
#| echo: false
#| message: false
#| warning: false
data |>
  summarise(
    mean_price = mean(Price, na.rm = TRUE),
    sd_price = sd(Price, na.rm = TRUE),
    min_price = min(Price, na.rm = TRUE),
    max_price = max(Price, na.rm = TRUE)
  ) |>
  kable(caption = "Summary Statistics of Stock Price")
```
@tbl-price shows the mean, standard deviation, minimum and maximum of the stock price between 2017/1/3 to 2026/3/27. 

## Transform Data
```{r}
#| label: fig-raw
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(Date, Price)) +
  geom_line() +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "red") +
  labs(title = "Stock Price Over Time", subtitle = "Dashed line = 2022-11-30")
```

```{r}
#| label: fig-rawacf
#| eval: true
#| echo: false
#| message: false
#| warning: false
acf(data$Price, main="ACF of Raw Stock Price")
```
@fig-rawacf indicates that the time series is not stationary, as the ACFs are out of the boundary and decay very slowly. Thus, we need to transform the data to make it stationary. For stock data, taking the natural log is a common way to transform the data.
```{r}
# log return the raw data
#| eval: true
#| echo: false
#| message: false
#| warning: false
data <- data |>
  mutate(log_price = log(Price),
         log_return = c(NA, diff(log_price)))

before_data <- data |> filter(Date < cutoff)
after_data  <- data |> filter(Date >= cutoff)
```

```{r}
#| label: fig-logreturn
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(x = Date, y = log_return)) +
  geom_line() +
  geom_vline(xintercept = cutoff, linetype = "dashed", color ="red") +
  labs(title = "Log Return Over Time",
       subtitle = "Dashed line = 2022-11-30",
       x = "Date", y = "Log Price")
```

```{r}
#| label: fig-logacf
#| eval: true
#| echo: false
#| message: false
#| warning: false
acf(na.omit(data$log_return), main="ACF of log return")
```
Figure @fig-logreturn shows the log return over time. The data fluctuates around zero, which suggests the mean is around zero. Figure @fig-logacf shows the ACF of log return. Most ACFs fall within the bound. Thus, the log return time series looks stationary.

## Compare log return
```{r}
#| label: tbl-log-return
#| tbl-cap: "Summary Statistics of Stock Price"
#| echo: false
#| message: false
#| warning: false
data |>
  mutate(period = ifelse(Date < cutoff, "Before", "After")) |>
  summarise(
    Mean_Return = mean(log_return, na.rm = TRUE),
    SD_Return = sd(log_return, na.rm = TRUE),
    .by = period
  ) |>
  mutate(
    Mean_Return = round(Mean_Return, 5),
    SD_Return = round(SD_Return, 5)
  ) |>
  kable()
```
According to table @tbl-log-return, the mean log return before and after the appearance of ChatGPT is both close to zero.

## Fit Model
```{r}
#| label: fig-pacf
#| eval: true
#| echo: false
#| message: false
#| warning: false
pacf(na.omit(data$log_return), main="PACF of log return")
```
According to figrue @fig-pacf, there is a strong pike at lag=1. So we fit AR(1), MA(1), and ARMA(1) to capture the time series.
```{r}
#AR(1)
AR1 = arima(data$log_return, order = c(1,0,0))
print(AR1)
```

```{r}
#MA(1)
MA1 = arima(data$log_return, order = c(0,0,1))
print(MA1)
```

```{r}
#ARMA(1,1)
ARMA1 = arima(data$log_return, order = c(1,0,1))
print(ARMA1)
```

## Forcasting
We are interested in how significant the data after 2022/11/30, when AI stared its prevalence, is to forecasting future stock market prices. We will being using data frames of different time sets: 2017/1/3 - 2026/3/27, 2017/1/3 - 2022/11/30 and 2022/11/30 - 2026/3/27. The forecast models we will be using are five "benchmark" methods: Mean method, Naive method, Seasonal naive method, (drift methods and decomposition method, note: will do this later). 

First, we aimed to examine whether AI has had an unexpected impact on the stock market and contributed to its fluctuations. We would like to use data from 2017/1/3 - 2022/11/30 to forecast data after 2022/11/30.

```{r, include=FALSE}
library(zoo)
library(lubridate) 
# [Link: Stack Overflow https://stackoverflow.com/questions/18755643/time-series-analysis-with-r-how-to-deal-with-daily-data]

# Create a zoo object
log_return_ts <- zoo(data$log_return, order.by = data$Date[-1])

# Plot the time series
plot(log_return_ts, main = "Daily Time Serie", xlab = "Date", ylab = "Log_return", type = "l", col = "blue")
```

```{r, include=FALSE}
x_train1 <- log_return_ts[index(log_return_ts) <= as.Date("2022-11-30")]
x_test1 <- log_return_ts[index(log_return_ts) > as.Date("2022-11-30")]

#forecasts
x_mean1 <- rep(mean(x_train1), length(x_test1)) #mean
x_naive1 <- rep(tail(x_train1,1), length(x_test1)) #naive
x_seasonal1 <- rep(c(tail(x_train1, 4)), length(x_test1)) #seasonal

# forecast errors
mean_errors1 <- x_test1 - x_mean1
naive_errors1 <- x_test1 - x_naive1
seasonal_errors1 <- x_test1 - x_seasonal1

# MAE(Scale-dependent Errors)
mean_MAE1 <- mean(abs(mean_errors1), na.rm = TRUE)
naive_MAE1 <- mean(abs(naive_errors1), na.rm = TRUE)
seasonal_MAE1 <- mean(abs(seasonal_errors1), na.rm = TRUE)

#RMSE
mean_RMSE1 <- sqrt(mean(mean_errors1^2, na.rm = TRUE))
naive_RMSE1 <- sqrt(mean(naive_errors1^2, na.rm = TRUE))
seasonal_RMSE1 <- sqrt(mean(seasonal_errors1^2, na.rm = TRUE))


```

```{r, include=FALSE}
cat("MAE - Mean Forecast:     ", mean_MAE1, "\n")
cat("MAE - Naive Forecast:    ", naive_MAE1, "\n")
cat("MAE - Seasonal Forecast: ", seasonal_MAE1, "\n\n")

cat("RMSE - Mean Forecast:     ", mean_RMSE1, "\n")
cat("RMSE - Naive Forecast:    ", naive_RMSE1, "\n")
cat("RMSE - Seasonal Forecast: ", seasonal_RMSE1, "\n")
```

```{r}
#| label: tbl-forecast_errors
#| eval: true
#| echo: false
#| message: false
#| warning: false

forecast_table <- data.frame(
  Method = c("Mean", "Naive", "Seasonal"),
  MAE    = c(mean_MAE1, naive_MAE1, seasonal_MAE1),
  RMSE   = c(mean_RMSE1, naive_RMSE1, seasonal_RMSE1)
)

kable(forecast_table, digits = 5, caption = "Forecast Errors")
```

According to @tbl-forecast_errors, all the scaled dependent errors are small, even for these easier approaches. It appears that the data prior to 2022/11/30 can effectively predict the data following the cutoff point. Furthermore, this suggests that AI have not yet contribute a significant growth in the stock market. As a result, we will use the entire time frame, allocating 80% for training and 20% for testing, to build a forecast model in predict prices in future 2026. 



