---
title: "main"
format: pdf
---
```{r, include=FALSE}
#packages
library("tidyverse")
library("ggplot2")
library("dplyr")
library("knitr")
```

# Section 1. Data Collection
The data are from 1/28/2019 to 3/28/2026. 


# Section 2. Variable Creation and Data Cleaning
Original dataset contains date as a character which may cause problems when using it as our time predictor, so we converted to the Date type in R. 
We observed outliers that are doubled the price of the remaining data points. Looking at the data, we found that for most weekends, prices are carried over from the Friday of the same week. These outliers correspond to weekend days, as indicated by their dates, and their volumes are recorded as empty strings. We decided to remove these outliers because they are not consistent with the general pattern, which they doubled the Friday price. 
Lastly, we split the dataset into two groups using November 30, 2022 (the release date of ChatGPT) as the cutoff. This is used for the comparison of the two time seires.

We will detrend the data (and difference) the data to remove the overall growth in the stock market and possible market cycles. 
```{r, include=FALSE}
#load data
data <- read.csv("XLK ETF Stock Price History.csv")
```

```{r, include=FALSE}
# change date 
data$Date <- as.Date(data$Date, format = "%m/%d/%Y") 
data <- data |>
  filter(trimws(Vol.) != "")
```

```{r, include=FALSE}
#split into two time frames
cutoff <- as.Date("2022-11-30")
before_data <- data[data$Date < cutoff, ]
after_data  <- data[data$Date >= cutoff, ]
```

# Section 3 Analytic Methods 
## regression model
### model assumption check

## detrend 
- check if data is white noise, or MA, or AR
```{r}
plot(data$Date, data$Price, type = "l", xlab="Date", ylab="Price", main = "Stock price over time (2019 - 2026)")
```

```{r}
par(mfrow = c(1,2))
plot(before_data$Date, before_data$Price, type = "l")
plot(after_data$Date, after_data$Price, type = "l", col = "blue")
```


```{r}
#detrend
model_data <- data.frame(
  t = 1:nrow(data),
  y = data$Price
)

fit <- lm(y ~ t, data = model_data)
data$detrended_price <- resid(fit)
acf(data$detrended_price)
```

## forecasting 

# Section 4
## EDA
```{r}
#| label: tbl-price
#| tbl-cap: "Summary Statistics of Stock Price"
#| echo: false
#| message: false
#| warning: false
data |>
  summarise(
    mean_price = mean(Price, na.rm = TRUE),
    sd_price = sd(Price, na.rm = TRUE),
    min_price = min(Price, na.rm = TRUE),
    max_price = max(Price, na.rm = TRUE)
  ) |>
  kable(caption = "Summary Statistics of Stock Price")
```
Table @tbl-price shows the mean, standard deviation, minimum and maximum of the stock price between 2017/1/3 to 2026/3/27. 

## Transform Data
```{r}
#| label: fig-raw
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(Date, Price)) +
  geom_line() +
  geom_vline(xintercept = cutoff, linetype = "dashed", color = "red") +
  labs(title = "Stock Price Over Time", subtitle = "Dashed line = 2022-11-30")
```

```{r}
#| label: fig-rawacf
#| eval: true
#| echo: false
#| message: false
#| warning: false
acf(data$Price, main="ACF of Raw Stock Price")
```
Figure @fig-rawacf indicates that the time series is not stationary, as the ACFs are out of the boundary and decay very slowly. Thus, we need to transform the data to make it stationary. For stock data, taking the natural log is a common way to transform the data.
```{r}
# log return the raw data
#| eval: true
#| echo: false
#| message: false
#| warning: false
data <- data |>
  mutate(log_price = log(Price),
         log_return = c(NA, diff(log_price)))

before_data <- data |> filter(Date < cutoff)
after_data  <- data |> filter(Date >= cutoff)
```

```{r}
#| label: fig-logreturn
#| eval: true
#| echo: false
#| message: false
#| warning: false
ggplot(data, aes(x = Date, y = log_return)) +
  geom_line() +
  geom_vline(xintercept = cutoff, linetype = "dashed", color ="red") +
  labs(title = "Log Return Over Time",
       subtitle = "Dashed line = 2022-11-30",
       x = "Date", y = "Log Price")
```

```{r}
#| label: fig-logacf
#| eval: true
#| echo: false
#| message: false
#| warning: false
acf(na.omit(data$log_return), main="ACF of log return")
```
Figure @fig-logreturn shows the log return over time. The data fluctuates around zero, which suggests the mean is around zero. Figure @fig-logacf shows the ACF of log return. Most ACFs fall within the bound. Thus, the log return time series looks stationary.

## Compare log return
```{r}
#| label: tbl-log-return
#| tbl-cap: "Summary Statistics of Stock Price"
#| echo: false
#| message: false
#| warning: false
data |>
  mutate(period = ifelse(Date < cutoff, "Before", "After")) |>
  summarise(
    Mean_Return = mean(log_return, na.rm = TRUE),
    SD_Return = sd(log_return, na.rm = TRUE),
    .by = period
  ) |>
  mutate(
    Mean_Return = round(Mean_Return, 5),
    SD_Return = round(SD_Return, 5)
  ) |>
  kable()
```
According to table @tbl-log-return, the mean log return before and after the appearance of ChatGPT is both close to zero.

## Fit Model
```{r}
#| label: fig-pacf
#| eval: true
#| echo: false
#| message: false
#| warning: false
pacf(na.omit(data$log_return), main="PACF of log return")
```
According to figrue @fig-pacf, there is a strong pike at lag=1. So we fit AR(1), MA(1), and ARMA(1) to capture the time series.
```{r}
#AR(1)
AR1 = arima(data$log_return, order = c(1,0,0))
print(AR1)
```

```{r}
#MA(1)
MA1 = arima(data$log_return, order = c(0,0,1))
print(MA1)
```

```{r}
#ARMA(1,1)
ARMA1 = arima(data$log_return, order = c(1,0,1))
print(ARMA1)
```

